Reconstructing the neutron-star equation of state with gravitational-wave detectors from a realistic population of inspiralling binary neutron stars

### 2014-09-16 paper.tex
Committing some changes before I pull Les' version

\begin{abstract}
Gravitational-wave observations of inspiralling binary neutron star systems can be used to measure the neutron-star equation of state (EOS) through the tidally induced shift in the waveform phase that depends on the tidal deformability parameter $\lambda$. Previous work has shown that $\lambda$, a function of the neutron-star EOS and mass, is marginally measurable by Advanced LIGO for a single event when including tidal information up to the merger frequency. In this work, we describe a method for stacking measurements of $\lambda$ from multiple inspiral events to measure the EOS. We use Markov chain Monte Carlo simulations to estimate the parameters of a 4-parameter piecewise polytrope EOS that matches theoretical EOS models to a few percent. We find that, for ``realistic'' event rates ($\sim 40$ binary neutron star inspiral events per year with signal-to-noise ratio $> 8$ in a single Advanced LIGO detector), combining a year of triple-coincident gravitational-wave data with the constraints from causality and recent high mass neutron-star measurements, the EOS above nuclear density can be measured to better than a factor of two in most cases. We also find that in the mass range $1M_\odot$--$2M_\odot$, the neutron-star radius can be measured to \ben{better than}{} $\pm 1$~km and the tidal deformability \ben{to better than}{can be measured to} $\pm 1 \times 10^{36}$~g~cm$^2$~s$^2$. The overwhelming majority of this information comes from the loudest $\sim 5$ events. Current uncertainties in the post-Newtonian waveform model, however, lead to systematic errors in the EOS measurement that are as large as the statistical errors, and more accurate waveform models are needed to minimize this error.
\end{abstract}

\pacs{
04.40.Dg, %Relativistic stars: structure, stability, and oscillations
26.60.Kp, %Equations of state of neutron-star matter
97.60.Jd  %Neutron stars
}

@@

[[Update the PACS numbers.]]

[[Eq 4 of Guillot et al. refers to Koranda et al. regarding a relation between the maximum radius and maximum mass. Is this the cause for the roughly straight line you get for the high-mass part of the prior?]]

@@

The measurability of tidal parameters in BNS systems with aLIGO for GW frequencies below 450Hz (prior to the last $\sim 20$ GW cycles before merger where higher order corrections to the tidal effect become important) was first examined for polytropic EOSs in~\cite{FlanaganHinderer2008} and for theoretical hadronic and quark matter EOSs in~\cite{HindererLackeyLangRead2010}.  These studies found that for a single aLIGO detector, tidal interactions were only observable during this early inspiral stage for stiff EOSs, NS masses below 1.4~$M_\odot$, and for \ben{rare}{}, nearby sources with signal-to-noise ratios (SNR) $\gtrsim 30$. Damour {\it et al.}, however, found that when also including EOS information from the last $\sim 20$ GW cycles prior to contact, tidal parameters are in fact observable \ben{even with more common}{ \les{for more likely}{}} SNRs of $\sim 16$~\cite{DamourNagarVillain2012}.

@@

In this paper we will demonstrate the advantages of parameterizing the EOS instead of $\lambda(M)$. The main benefit comes from including prior knowledge of the EOS that would be more difficult to incorporate into the $\lambda(M)$ fit. As will be discussed in Section~\ref{sec:eos}, the requirements that the EOS is monotonic and causal provide the simple constraint that the slope of the EOS is in the range $0 \le dp/d\epsilon \le c^2$, where $\epsilon$ is the energy density \les{and $c$ is the speed of light}{}. Additionally, the constraints from the observed $2M_\odot$ NSs can be included by simply rejecting any EOS parameters that lead to a NS maximum mass below $2M_\odot$. We will find that incorporating this information allows us to make significantly stronger statements about the EOS, radius, and tidal deformability, and this is true for a narrower, more realistic range of NS masses than simulated by Del Pozzo {\it et al.}~\cite{DelPozzoLiAgathos2013}.

The BNS inspiral signal is almost exactly characterized by the mass, spin, and tidal deformability of each NS, and is therefore free from the intrinsic variability and uncertainties that bias electromagnetic measurements of the mass and radius. Currently unknown terms in the post-Newtonian (PN) waveform model, however, will lead to systematic errors in recovering the tidal deformabilities \ben{}{and, to a lesser extent, masses}. Using the Fisher matrix approximation~\cite{Favata2014, YagiYunes2014} as well as a Markov chain Monte Carlo (MCMC) Bayesian analysis~\cite{WadeCreightonOchsner2014}, it was found that the systematic error in $\lambda$ from waveform uncertainties can be as large as the statistical error for a single BNS inspiral observation. When stacking observations the problem becomes worse, as the statistical errors decrease with more observations but the systematic errors do not, and we will discuss how systematic errors impact the recovery of the EOS.

@@

\ben{Because the EOS can be calculated from the}{Because the EOS can be found from a full knowledge of the} $R(M)$~\cite{Lindblom1992} or $\lambda(M)$ relations~\cite{LindblomIndik2012, LindblomIndik2014}, we can choose to parameterize either $R(M)$, $\lambda(M)$ or $p(\rho)$. As Del Pozzo {\it et al.}\ found by parameterizing $\lambda(M)$ with a linear fit, it is difficult to accurately measure $\lambda(M)$ over a wide range of masses with BNS inspiral observations. Instead they found that $\lambda$ could only be accurately measured for a specific fiducial mass which they chose to be $1.4M_\odot$. This is because, when parameterizing the function $\lambda(M)$, there is little {\it a priori} information about the allowed functional form of $\lambda(M)$. Lattimer and Steiner found similar results for mass and radius observations in Section~4.1 of Ref.~\cite{LattimerSteiner2014} when $R(M)$ was parameterized instead of the EOS. In contrast, it is much simpler to place useful constraints on the functional form of the EOS fit, and this allows one to make significantly stronger statements about the EOS and the behavior of the $\lambda(M)$ and $R(M)$ curves.

@@

The {\it a priori} constraints on the EOS that we will use are:
\begin{enumerate}
\item The EOS can be considered known below a density $\rho_0$ defined below, and we use a fixed EOS below that density.
\item The EOS must be a monotonically increasing function $dp/d\epsilon \ge 0$, where $\epsilon$ is the energy density, to satisfy thermodynamic stability.
\item The speed of sound $v_s = \sqrt{dp/d\epsilon}$ must be less than the speed of light, ensuring causality. 
\item The EOS must allow for a maximum NS mass greater than observed masses. Recent pulsar mass measurements in two neutron star-white dwarf binaries have provided convincing evidence that the maximum NS mass is $\gtrsim 2M_\odot$. The pulsar J1614-2230 was found to have a mass of $1.97\pm0.04M_\odot$ ($1\sigma$ confidence)~\cite{DemorestPennucciRansom2010}, and the pulsar J0348+0432 was found to have a mass of $2.01\pm0.04M_\odot$ ($1\sigma$ confidence)~\cite{AntoniadisFreireWex2013}. We will take the $2\sigma$ lower bound of $1.93M_\odot$ on the mass of J0348+0432 as a solid lower bound on the maximum NS mass.
\end{enumerate}

Although other constraints exist, such as the constraint that the maximum mass-shedding (Kepler) frequency be greater than observed spin frequencies, these turn out to be less useful, so we will focus on the ones listed above. See Ref.~\cite{ReadLackey2009} for a discussion of other constraints.

@@

For each piece of a piecewise polytrope, the pressure $p$ in the rest-mass density interval $\rho_{i-1} < \rho < \rho_i$ is defined by
\begin{equation}
p(\rho) = K_i \rho^{\Gamma_i},
\end{equation}
where $\Gamma_i$ is the adiabatic index and the constant $K_i$ is chosen such that $p$ is continuous at the boundary $\rho_{i-1}$. For the core of the star, we use 3 polytropes with adiabatic indices $\Gamma_1$, $\Gamma_2$, and $\Gamma_3$, separated by the fixed dividing densities $\rho_1 = 10^{14.7}$~g/cm$^3$ and $\rho_2 = 10^{15}$~g/cm$^3$. These fixed dividing densities were chosen to minimize the least-squares error between the piecewise polytrope fit and a set of 34 tabulated theoretical EOSs~\cite{ReadLackey2009}. In addition to the three adiabatic indices, we also require an additional parameter to determine the overall pressure scaling which we choose to be $p_1 = p(\rho_1)$, the pressure at the first dividing density. We will therefore use as our four EOS parameters $\vec E = \{\log(p_1), \Gamma_1, \Gamma_2, \Gamma_3\}$. \les{($E$ was used as energy in Section II.  We should change one or the other, I think)}{} \ben{[[I think it's clear from context and the fact that the parameters $\vec E$ are always a vector. Was this not true?]]}{}

@@

\begin{table*}[!htb]
\begin{center}
\caption{Comparison between tabulated EOSs and their best fits. The parameters that provide the best fit to the candidate EOSs as well as the residual are given. $p_1$ is in units of dyne/cm$^2$. Values for observables calculated using the tabulated EOSs are also given. $v_{\rm s,max}$ is the maximum speed of sound below $\rho_{\rm c, max}$. $M_{\rm max}$ is the maximum nonrotating mass configuration in units of $M_\odot$. $R_{1.4}$ is the radius of a $1.4~M_\odot$ star in units of km. $\Lambda_{1.4}$ is the tidal deformability of a $1.4~M_\odot$ star. The difference in calculated values for each observable when using the tabulated EOS $(O_{\rm tab})$ versus the best fit parameterized EOS $(O_{\rm fit})$ is calculated with $(O_{\rm fit}/O_{\rm tab}-1)100$. \red{[[Also include central density at 1.4 and maximum mass?]]} }
\begin{tabular}{l|ccccl|rr|rr|rr|rr}
\hline\hline
EOS & $\log(p_1)$ & $\Gamma_1$ & $\Gamma_2$ & $\Gamma_3$ & residual &
$v_{s, \rm max}$ & \% & $M_{\rm max}$ & \% & $R_{1.4}$ & \% & $\Lambda_{1.4}$ & \% \\
\hline
SLy&34.384&3.005&2.988&2.851&0.0020& 0.989&1.41& 2.049&0.02& 11.736&-0.21\\
ENG&34.437&3.514&3.130&3.168&0.015& 1.000&10.71& 2.240&-0.05& 12.059&-0.69\\
MPA1&34.495&3.446&3.572&2.887&0.0081& 0.994&4.91& 2.461&-0.16& 12.473&-0.26\\
MS1&34.858&3.224&3.033&1.325&0.019& 0.888&12.44& 2.767&-0.54& 14.918&0.06\\
MS1b&34.855&3.456&3.011&1.425&0.015& 0.889&11.38& 2.776&-1.03& 14.583&-0.32\\
H4&34.669&2.909&2.246&2.144&0.0028& 0.685&4.52& 2.032&-0.85& 13.774&1.34\\
ALF2&34.616&4.070&2.411&1.890&0.043& 0.642&1.50& 2.086&-5.26& 13.188&-3.66\\
\hline\hline
\end{tabular}
\label{tab:eosfit}
\end{center}
\end{table*}

>>>>

\begin{abstract}
Gravitational-wave observations of inspiralling binary neutron star systems can be used to measure the neutron-star equation of state (EOS) through the tidally induced shift in the waveform phase that depends on the tidal deformability parameter $\lambda$. Previous work has shown that $\lambda$, a function of the neutron-star EOS and mass, is marginally measurable by Advanced LIGO for a single event when including tidal information up to the merger frequency. In this work, we describe a method for stacking measurements of $\lambda$ from multiple inspiral events to measure the EOS. We use Markov chain Monte Carlo simulations to estimate the parameters of a 4-parameter piecewise polytrope EOS that matches theoretical EOS models to a few percent. We find that, for ``realistic'' event rates ($\sim 40$ binary neutron star inspiral events per year with signal-to-noise ratio $> 8$ in a single Advanced LIGO detector), combining a year of triple-coincident gravitational-wave data with the constraints from causality and recent high mass neutron-star measurements, the EOS above nuclear density can be measured to better than a factor of two in most cases. We also find that in the mass range $1M_\odot$--$2M_\odot$, the neutron-star radius can be measured to better than $\pm 1$~km and the tidal deformability to better than $\pm 1 \times 10^{36}$~g~cm$^2$~s$^2$. The overwhelming majority of this information comes from the loudest $\sim 5$ events. Current uncertainties in the post-Newtonian waveform model, however, lead to systematic errors in the EOS measurement that are as large as the statistical errors, and more accurate waveform models are needed to minimize this error.
\end{abstract}

\pacs{
97.60.Jd  %Neutron stars
26.60.Kp, %Equations of state of neutron-star matter
04.30.Tv, %Gravitational-wave astrophysics
}

@@

%[[Eq 4 of Guillot et al. refers to Koranda et al. regarding a relation between the maximum radius and maximum mass. Is this the cause for the roughly straight line you get for the high-mass part of the prior?]]

@@

The measurability of tidal parameters in BNS systems with aLIGO for GW frequencies below 450Hz (prior to the last $\sim 20$ GW cycles before merger where higher order corrections to the tidal effect become important) was first examined for polytropic EOSs in~\cite{FlanaganHinderer2008} and for theoretical hadronic and quark matter EOSs in~\cite{HindererLackeyLangRead2010}.  These studies found that for a single aLIGO detector, tidal interactions were only observable during this early inspiral stage for stiff EOSs, NS masses below 1.4~$M_\odot$, and for rare, nearby sources with signal-to-noise ratios (SNR) $\gtrsim 30$. Damour {\it et al.}, however, found that when also including EOS information from the last $\sim 20$ GW cycles prior to contact, tidal parameters are in fact observable even with more common SNRs of $\sim 16$~\cite{DamourNagarVillain2012}.

@@

In this paper we will demonstrate the advantages of parameterizing the EOS instead of $\lambda(M)$. The main benefit comes from including prior knowledge of the EOS that would be more difficult to incorporate into the $\lambda(M)$ fit. As will be discussed in Section~\ref{sec:eos}, the requirements that the EOS is monotonic and causal provide the simple constraint that the slope of the EOS is in the range $0 \le dp/d\epsilon \le c^2$, where $\epsilon$ is the energy density and $c$ is the speed of light. Additionally, the constraints from the observed $2M_\odot$ NSs can be included by simply rejecting any EOS parameters that lead to a NS maximum mass below $2M_\odot$. We will find that incorporating this information allows us to make significantly stronger statements about the EOS, radius, and tidal deformability, and this is true for a narrower, more realistic range of NS masses than simulated by Del Pozzo {\it et al.}~\cite{DelPozzoLiAgathos2013}.

The BNS inspiral signal is almost exactly characterized by the mass, spin, and tidal deformability of each NS, and is therefore free from the intrinsic variability and uncertainties that bias electromagnetic measurements of the mass and radius. Currently unknown terms in the post-Newtonian (PN) waveform model, however, will lead to systematic errors in recovering the tidal deformabilities. Using the Fisher matrix approximation~\cite{Favata2014, YagiYunes2014} as well as a Markov chain Monte Carlo (MCMC) Bayesian analysis~\cite{WadeCreightonOchsner2014}, it was found that the systematic error in $\lambda$ from waveform uncertainties can be as large as the statistical error for a single BNS inspiral observation. When stacking observations the problem becomes worse, as the statistical errors decrease with more observations but the systematic errors do not, and we will discuss how systematic errors impact the recovery of the EOS.

@@

Because the EOS can be calculated from the $R(M)$~\cite{Lindblom1992} or $\lambda(M)$ relations~\cite{LindblomIndik2012, LindblomIndik2014}, we can choose to parameterize either $R(M)$, $\lambda(M)$ or $p(\rho)$ if we want to reconstruct the EOS. As Del Pozzo {\it et al.}\ found by parameterizing $\lambda(M)$ with a linear fit, it is difficult to accurately measure $\lambda(M)$ over a wide range of masses with BNS inspiral observations. Instead, they found that $\lambda$ could only be accurately measured for a specific fiducial mass which they chose to be $1.4M_\odot$. This is because, when parameterizing the function $\lambda(M)$, there is little {\it a priori} information about the allowed functional form of $\lambda(M)$. Lattimer and Steiner found similar results for mass and radius observations (Section~4.1 of Ref.~\cite{LattimerSteiner2014}) when they parameterized $R(M)$ instead of the EOS. In contrast, it is much simpler to place useful constraints on the functional form of the EOS fit, and this allows one to make significantly stronger statements about the EOS and the behavior of the $\lambda(M)$ and $R(M)$ curves.

@@

The {\it a priori} constraints on the EOS that we will use are:
\begin{enumerate}
\item The EOS can be considered known below a certain density $\rho_0$, and we use a fixed EOS below that density.
\item The EOS must be a monotonically increasing function $dp/d\epsilon \ge 0$, where $\epsilon$ is the energy density, to satisfy thermodynamic stability.
\item The speed of sound $v_s = \sqrt{dp/d\epsilon}$ must be less than the speed of light, ensuring causality. 
\item The EOS must allow for a maximum NS mass greater than observed masses. Recent pulsar mass measurements in two neutron star-white dwarf binaries have provided convincing evidence that the maximum NS mass is $\gtrsim 2M_\odot$. The pulsar J1614-2230 was found to have a mass of $1.97\pm0.04M_\odot$ ($1\sigma$ confidence)~\cite{DemorestPennucciRansom2010}, and the pulsar J0348+0432 was found to have a mass of $2.01\pm0.04M_\odot$ ($1\sigma$ confidence)~\cite{AntoniadisFreireWex2013}. We will take the $2\sigma$ lower bound of $1.93M_\odot$ on the mass of J0348+0432 as a solid lower bound on the maximum NS mass.
\end{enumerate}
Although other constraints exist, such as the constraint that the maximum mass-shedding (Kepler) frequency be greater than observed spin frequencies, these turn out to be less useful, so we will focus on the ones listed above. See Ref.~\cite{ReadLackey2009} for a discussion of other constraints.

@@

For each piece of a piecewise polytrope, the pressure $p$ in the rest-mass density interval $\rho_{i-1} < \rho < \rho_i$ is defined by
\begin{equation}
p(\rho) = K_i \rho^{\Gamma_i},
\end{equation}
where $\Gamma_i$ is the adiabatic index and the constant $K_i$ is chosen such that $p$ is continuous at the boundary $\rho_{i-1}$. For the core of the star, we use 3 polytropes with adiabatic indices $\Gamma_1$, $\Gamma_2$, and $\Gamma_3$ separated by the fixed dividing densities $\rho_1 = 10^{14.7}$~g/cm$^3$ and $\rho_2 = 10^{15}$~g/cm$^3$. These fixed dividing densities were chosen to minimize the least-squares error between the piecewise polytrope fit and a set of 34 tabulated theoretical EOSs~\cite{ReadLackey2009}. In addition to the three adiabatic indices, we also require an additional parameter to determine the overall pressure scaling which we choose to be $p_1 = p(\rho_1)$, the pressure at the first dividing density. We will therefore use as our four EOS parameters $\vec E = \{\log(p_1), \Gamma_1, \Gamma_2, \Gamma_3\}$.

@@

\begin{table*}[!htb]
\begin{center}
\caption{Comparison between tabulated EOSs and their best fits. The parameters $\log(p_1/\text{(dyne~cm$^{-2}$)})$, $\Gamma_1$, $\Gamma_2$, and $\Gamma_3$ are the values that minimize the least-squares residual. Observables for the tabulated EOSs are also shown; $v_{\rm s,max}$ ($c$) is the maximum speed of sound below $\rho_{\rm c, max}$, $M_{\rm max}$ ($M_\odot$) is the maximum mass, $R_{1.4}$ (km) is the radius of a $1.4~M_\odot$ NS, and $\lambda_{1.4}$ ($10^{36}$~g~cm$^2$~s$^2$) is the tidal deformability of a $1.4~M_\odot$ NS. The percent error for each observable when using the tabulated EOS $(O_{\rm tab})$ versus the best-fit parameterized EOS $(O_{\rm fit})$ is also given by $(O_{\rm fit}/O_{\rm tab}-1)100$.}
\begin{tabular}{l|ccccl|rr|rr|rr|rr}
\hline\hline
EOS & $\log(p_1)$ & $\Gamma_1$ & $\Gamma_2$ & $\Gamma_3$ & residual &
$v_{s, \rm max}$ & \% & $M_{\rm max}$ & \% & $R_{1.4}$ & \% & $\lambda_{1.4}$ & \% \\
\hline
SLy&34.384&3.005&2.988&2.851&0.0020& 0.989&1.41& 2.049&0.02& 11.736&-0.21\\
ENG&34.437&3.514&3.130&3.168&0.015& 1.000&10.71& 2.240&-0.05& 12.059&-0.69\\
MPA1&34.495&3.446&3.572&2.887&0.0081& 0.994&4.91& 2.461&-0.16& 12.473&-0.26\\
MS1&34.858&3.224&3.033&1.325&0.019& 0.888&12.44& 2.767&-0.54& 14.918&0.06\\
MS1b&34.855&3.456&3.011&1.425&0.015& 0.889&11.38& 2.776&-1.03& 14.583&-0.32\\
H4&34.669&2.909&2.246&2.144&0.0028& 0.685&4.52& 2.032&-0.85& 13.774&1.34\\
ALF2&34.616&4.070&2.411&1.890&0.043& 0.642&1.50& 2.086&-5.26& 13.188&-3.66\\
\hline\hline
\end{tabular}
\label{tab:eosfit}
\end{center}
\end{table*}
